On the propriety of restricted Boltzmann machines

### 2016-11-23 writing/draft.Rmd
last few vardeman edits

---
title: "On the propriety of restricted Boltzmann machines"
author: | 
  | Andee Kaplan, Daniel Nordman, and Stephen Vardeman
  | Iowa State University
  | ajkaplan@iastate.edu, dnordman@iastate.edu, vardeman@iastate.edu
date: "November, 2016"
linestretch: 2 
output: 
  pdf_document:
    fig_caption: yes
    includes:
      in_header: latex/header.tex
    number_sections: yes
bibliography: references/refs.bib
abstract: |
  A restricted Boltzmann machine (RBM) is an undirected graphical model constructed for discrete or continuous random variables, with two layers, one hidden and one visible, and no conditional dependency within a layer. In recent years, RBMs have risen to prominence due to their connection to deep learning. By treating a hidden layer of one RBM as the visible layer in a second RBM, a deep architecture can be created. RBMs are thought to thereby have the ability to encode very complex and rich structures in data, making them attractive for supervised learning. However, the generative behavior of RBMs is largely unexplored. In this paper, we discuss the relationship between RBM parameter specification in the binary case and the prevalence of undesirable model properties such as degeneracy, instability and uninterpretability. We also describe the difficulties that arise in likelihood-based and Bayes fitting of such (highly flexible) models, especially as Gibbs sampling (quasi-Bayes) methods are often advocated for the RBM model structure.
---

@@

where $\boldsymbol \theta \equiv (\theta_{11}, \dots, \theta_{1\nh}, \dots, \theta_{\nv 1}, \dots, \theta_{\nv \nh}, \theta_{v_1}, \dots, \theta_{v_\nv}, \theta_{h_1}, \dots, \theta_{h_\nh}) \in \mathbb{R}^{\nv + \nh + \nv*\nh}$ denotes the vector of model parameters and the denominator $\gamma(\boldsymbol \theta) = \sum_{\boldsymbol x \in \mathcal{C}^{\nh+\nv}}\exp\left(\sum_{i = 1}^\nv \sum_{j=1}^\nh \theta_{ij} v_i h_j + \sum_{i = 1}^\nv\theta_{v_i} v_i + \sum_{j = 1}^\nh\theta_{h_j} h_j\right)$ is the normalizing function that ensures the probabilities (\ref{eqn:pmf}) sum to one. For $\boldsymbol x = (h_1, \dots, h_\nh, v_1, \dots, v_\nh) \in \mathcal{C}^{\nv + \nh}$ and 
\begin{align}
\boldsymbol t(\boldsymbol x) &= (h_1, \dots, h_\nh, v_1, \dots, v_\nv, v_1 h_1, \dots, v_\nv h_\nh) \in \mathcal{C}^{\nh + \nv + \nh*\nv}, \label{eqn:t}
\end{align}
let $\mathcal{T} = \{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{C}^{\nv + \nh}\} \subset \mathbb{R}^{\nv + \nh + \nv * \nh}$ be the set of possible values for the vector of variables needed to compute probabilities (\ref{eqn:pmf}) in the model and write $Q_{\boldsymbol \theta}(\boldsymbol x) = \sum\limits_{i = 1}^\nh\sum\limits_{j=1}^\nv \theta_{ij} h_i v_j + \sum\limits_{i = 1}^\nh\theta_{hi} h_i + \sum\limits_{j = 1}^\nv\theta_{vj} v_j$ for the "neg-potential" function. The RBM model is parameterized by $\boldsymbol \theta$ containing two types of parameters, main effects and interaction effects. The main effects parameters ($\{\theta_{v_i}, \theta_{h_j}\}_{\substack{i = 1, \dots, \nv,\\j = 1, \dots, \nh}}$) weight the values of the visible $v_i$ and hidden $h_j$ nodes in probabilities (\ref{eqn:pmf}) and the interaction effects parameters ($\theta_{ij}$) weight the values of the connections $v_i h_j$, or dependencies, between hidden and visible layers.

Due to the potential size of the model, the normalizing constant $\gamma(\boldsymbol \theta)$ can be practically impossible to calculate, making simple estimation of the model parameter vector problematic. A kind of Gibbs sampling can be tried (due to conditional architecture of the RBM, i.e. visibles given hiddens or vice versa). Specifically, the conditional independence of nodes in each layer, given those nodes in the other layer, allows for stepwise simulation of both hidden layers and model parameters possible (e.g., contrastive divergence of @hinton2002training or Bayes methods in Section \ref{model-fitting}).

@@

## Degeneracy, instability, and uninterpretability... Oh my!
The highly flexible nature of a RBM (having as it does $\nh + \nv + \nh*\nv$ parameters) creates at least three kinds of potential model impropriety of concern that we will call *degeneracy*, *instability*, and *uninterpretability*. In this section we define these characteristics, consider how to detect them in a RBM, and point out relationships among them.

### Near-degeneracy
In Random Graph Model theory, *model degeneracy* means there is a disproportionate amount of probability placed on only a few elements of the sample space, $\mathcal{X}$, by the model [@handcock2003assessing]. For random graph models, $\mathcal{X}$ denotes all possible graphs that can be constructed from a set of nodes and an exponentially parameterized random graph model has a distribution of the form 
\begin{align*}
f_{\boldsymbol \theta} (\boldsymbol x) = \frac{\exp\left(\boldsymbol \theta^T \boldsymbol t(\boldsymbol x)\right)}{\gamma(\boldsymbol \theta)}, \boldsymbol x \in \mathcal{X},
\end{align*}
where $\boldsymbol \theta \in \Theta \subset \mathbb{R}^q$ is the model parameter, and $\boldsymbol t: \mathcal{X} \rightarrow \mathbb{R}^q$ is a vector of statistics based on the adjacency matrix of a graph. Here, as earlier, $\gamma(\boldsymbol \theta) = \sum_{\boldsymbol x \in \mathcal{X}} \exp\left(\boldsymbol \theta^T \boldsymbol t(\boldsymbol x)\right)$ is the normalizing function. Let $C$ denote the convex hull of the potential outcomes of sufficient statistics, $\{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{X}\}$. @handcock2003assessing classify an exponentially parametrized random graph model at $\boldsymbol \theta$ as *near-degenerate* if the mean value of the vector of sufficient statistics under $\boldsymbol \theta$, $\boldsymbol \mu(\boldsymbol \theta) = \text{E}_{\boldsymbol \theta}\boldsymbol t( \boldsymbol X)$, is close to the boundary of $C$. This makes sense because if a model is near-degenerate, in the sense that only a small number of elements of the sample space $\mathcal{X}$ have positive probability, the expected value $\text{E}_{\boldsymbol \theta}\boldsymbol t( \boldsymbol X)$ is an average of that same small number of values of $t( \boldsymbol x)$ and can be expected to not be pulled deep into the interior of $C$.

@@

Considering a class of exponential families of distributions, @schweinberger2011instability introduced a concept of model defficiency related to *instability*. Instability can be roughly thought of as excessive sensitivity in the model, where small changes in the components of potential data outcomes, $\boldsymbol x$, may lead to substantial changes in the probability function $f_{\boldsymbol \theta}(\boldsymbol x)$. To quantify "instability" more rigorously (particularly beyond the definition given by @schweinberger2011instability) it is useful to consider how RBM models might be expanded to incorporate more and more visibles. When increasing the size of RBM models, it becomes necessary to grow the number of model parameters (and in this process one may also arbitrarily expand the number of hidden variables used). To this end, let $\boldsymbol \theta_{\nv} \equiv (\theta_{v_1}, \dots, \theta_{v_\nv}, \theta_{h_1}, \dots, \theta_{h_\nh}, \theta_{11}, \dots, \theta_{\nv \nh}), \nv \ge 1$, denote an element of a sequence of RBM parameters indexed by the number $\nv$ of visibles $(V_1, \dots, V_\nv)$ and define a (scaled) extremal log-probability ratio of the RBM model at $\boldsymbol \theta_{\nv}$ as
\begin{align}
\frac{1}{\nv} \log \left[\frac{\max\limits_{(v_1, \dots, v_\nv) \in \mathcal{C}^\nv}P_{\boldsymbol \theta_\nv}(v_1, \dots, v_\nv)}{\min\limits_{(v_1, \dots, v_\nv) \in \mathcal{C}^\nv}P_{\boldsymbol \theta_\nv}(v_1, \dots, v_\nv)}\right] \equiv \frac{1}{\nv} \text{ELPR}(\boldsymbol \theta_\nv) \label{eqn:elpr}
\end{align}
where $P_{\boldsymbol \theta_\nv}(v_1, \dots, v_\nv) \propto \sum_{\{h_j\}\in \mathcal{C}}\exp\left(\sum_{i = 1}^\nv \theta_{v_i} v_i + \sum_{j = 1}^\nh \theta_{h_j} h_j + \sum_{i = 1}^\nv \sum_{j = 1}^\nh \theta_{ij} v_i h_j\right)$ is the RBM probability of observing outcome $(v_1, \dots, v_\nv)$ for the visible variables $(V_1, \dots, V_\nv)$ under parameter vector $\boldsymbol \theta_\nv$.

@@

S-unstable RBM model sequences are undesirable for several reasons. One is that, as mentioned above, small changes in data can lead to overly-sensitive changes in probability. Consider, for example, 
$$
\Delta(\boldsymbol \theta_\nv) \equiv \max \left\{\log \frac{P_{\boldsymbol \theta_\nv}(v_1, \dots, v_\nv)}{P_{\boldsymbol \theta_\nv}(v_1^*, \dots, v_\nv^*)} : (v_1, \dots, v_\nv) \& (v_1^*, \dots, v_\nv^*) \in \mathcal{C}^\nv \text{ differ by exactly one component}\right\},
$$
denoting the biggest log-probability ratio for a one component change in data outcomes (visibles) at a RBM parameter $\boldsymbol \theta_\nv$. We then have the following result.

@@

\begin{prop}
\label{prop:degen}
For an S-unstable RBM model sequence and any given $0 < \epsilon < 1$, $P_{\boldsymbol \theta_\nv}\left((v_1, \dots, v_\nv) \in M_{\epsilon, \boldsymbol \theta_\nv}\right) \rightarrow 1$ as $\nv \rightarrow \infty$.
\end{prop}

In other words, S-unstable RBM model sequences are gwauranteed to stack up all probability on arbitrarily narrow outcomes for visibles. Proofs of Propositions \ref{prop:instab}-\ref{prop:degen} appear in @kaplan2016note. These findings also have counterparts to results in @schweinberger2011instability, but, unlike results there, we do not limit consideration to exponential family forms with a fixed number of parameters.

### Uninterpretability
For spatial Markov models, @kaiser2007statistical defines a measure of model impropriety he calls *uninterpretability*, which is characterized by dependence parameters in a model being so extreme that marginal mean-structures fail to hold as anticipated by a model statement. We adapt this notion to RBM models as follows. Note that in a RBM, the parameters $\theta_{v_1}, \dots, \theta_{v_\nv}$ and $\theta_{h_1}, \dots, \theta_{h_\nh}$ are naturally associated with main effects of visible and hidden variables and can be interpreted as (logit functions of) means for variables $V_1, \dots, V_\nv, H_1, \dots, H_\nh$ in a model with no interaction parameters, $\theta_{ij} = 0, i = 1, \dots, \nv, j = 1, \dots, \nh$. That is, with no interaction parameters, we have from (\ref{eqn:pmf}) that
$$
P_\theta(V_i=1) \propto e^{\theta_{v_i}} \quad \text{ and } \quad P_\theta(H_j=1) \propto e^{\theta_{h_j}}, \quad  i=1,\ldots,\nv,j=1,\ldots,\nh
$$
so that, for example, $logit(P_\theta(V_i=1))=\theta_{v_i}$ (or $2 \theta_{v_i}$) under the coding $\mathcal{C}=\{0,1\}$ (or $\{-1,1\}$). Hence, these main effect parameters have a clear interpretation under an independence model (one with $\theta_{ij} = 0$) but this interpretation can break down as interaction parameters increase in magnitude relative to the size of the main effects. In such cases, the main effect parameters $\theta_{v_1}$ and $\theta_{h_j}$ are no longer interpretable as originally intended in the models (statements of marginal means) and the dependence parameters are so large as to dominate the entire model probability structure (also destroying the model interpretation of dependence as local conditional modifications of an overall marginal mean structure). To assess which parameter values $\boldsymbol \theta$ may cause difficulties in interpretation, we use the difference $\text{E}\left[\boldsymbol X | \boldsymbol \theta\right] - \text{E}\left[\boldsymbol X | \boldsymbol \theta^* \right]$ between two model expectations: E$\left[\boldsymbol X | \boldsymbol \theta\right]$ at $\boldsymbol \theta$ and expectations E$\left[\boldsymbol X | \boldsymbol \theta^* \right]$ given independence where $\boldsymbol \theta^*$ matches $\boldsymbol \theta$ for all main effects but otherwise has $\theta_{ij} = 0$ for $i = 1, \dots, \nv, j = 1, \dots, \nh$. Hence, $\boldsymbol \theta^*$ and $\boldsymbol \theta$ have the same main effects but $\boldsymbol \theta^*$ has $0$ dependence parameters. Uninterpretability is then avoided at a parametric specification $\boldsymbol \theta$ if the model expected value at $\boldsymbol \theta$ is not very different from the corresponding model expectation under independence. Using this, it is possible to investigate what parametric conditions lead to uninterpretability in a model versus those that guarantee interpretable models. If $||\text{E}\left[\boldsymbol X | \boldsymbol \theta\right] - \text{E}\left[\boldsymbol X | \boldsymbol \theta^*\right]||$ is large, then the RBM model with parameter vector $\boldsymbol \theta$ is said to be uninterpetable. The quantities to compare in the RBM case are

@@

We next explore and numerically explain the relationship between values of $\boldsymbol \theta$ and the three measures of model impropriety, near-degeneracy, instability, and uninterpretability, for RBM models of varying sizes.

@@

Figure \ref{fig:uninterp} shows that the absolute difference between coordinates of the vector of model expectations, E$\left[\boldsymbol X | \boldsymbol \theta\right]$ and corresponding expectations E$\left[\boldsymbol X | \boldsymbol \theta^*\right ]$ given independence grow for the toy RBM as the values of $\boldsymbol \theta$ are farther from zero, especially for large magnitudes of $\theta_{11}$. This is a third indication supporting that parameter vectors of large magnitude lead to model impropriety in a RBM.

@@

In addition to the argument that the $-1/1$ data encoding mitigates some perceived prevalence of near-degeneracy it also has the benefit of providing a guaranteed-to-be non-degenerate model at $\boldsymbol \theta = \boldsymbol 0 \in \mathbb{R}^{\nh + \nv + \nh*\nv}$, where the zero vector that serves as the natural center of the parameter space and induces the simplest possible model properties for the RBM (i.e., at $\boldsymbol \theta = \boldsymbol 0$, all variables are independent and visible variables are independent and uniformly distributed on $\{-1,1\}^\nv$). The proof of this and further exploration of the equivalence of the $\boldsymbol \theta$ parameterization of the RBM model class and parameterization by $\boldsymbol \mu(\boldsymbol \theta)$ is in the supplementary materials. Hence while from some computing perspectives $0/1$ coding might seem most natural, the $-1/1$ coding is far more convenient and interpretable from the point of view of statistical modeling, where it makes parameters simply interpreted in terms of symmetrically defined main effects and interactions. In light of all of these matters we will henceforth employ the $-1/1$ coding.

@@

To explore the impact of RBM parameter vector $\boldsymbol \theta$ magnitude on near-degeneracy, instability, and uninterpretability, we consider models of small size. For $\nh, \nv \in \{1, \dots, 4\}$, we sample `r dim(res$samp[[1]])[1]` values of $\boldsymbol \theta$ with various magnitudes (details to follow). For each set of parameters we then calculate metrics of model impropriety introduced in Section \ref{degeneracy-instability-and-uninterpretability-oh-my} based on $\boldsymbol \mu(\boldsymbol \theta)$, $\text{ELPR}(\boldsymbol \theta)/\nv$, and the coordinates of $\left|\text{E}\left[\boldsymbol X | \boldsymbol \theta\right] -   \text{E}\left[\boldsymbol X | \boldsymbol \theta^* \right] \right|$. In the case of near-degeneracy, we classify each model as near-degenerate or "viable" based on the distance of $\boldsymbol \mu(\boldsymbol \theta)$ from the boundary of the convex hull of $\mathcal{T}$ and look at the fraction of models that are "near-degenerate," meaning they are within a small distance $\epsilon > 0$ of the boundary of the convex hull. We define a "small" through a rough estimation of the volume of the hull for each model size. We pick $\epsilon_0 = 0.05$ for $\nh=\nv=1$ and then, for every other $\nh$ and $\nv$, set $m=\nh+\nv+\nv*\nh$ and pick $\epsilon$ so that $1-(1-2\epsilon_0)^3 = 1 - (1-2\epsilon)^m$. In this way, we roughly scale the volume of the "small distance" to the boundary of the convex hull to be equivalent across model dimensions.

>>>>

---
title: "On the propriety of restricted Boltzmann machines"
author: | 
  | Andee Kaplan, Daniel Nordman, and Stephen Vardeman
  | Iowa State University
  | ajkaplan@iastate.edu, dnordman@iastate.edu, vardeman@iastate.edu
date: "November, 2016"
linestretch: 2 
output: 
  pdf_document:
    fig_caption: yes
    includes:
      in_header: latex/header.tex
    number_sections: yes
bibliography: references/refs.bib
abstract: |
  A restricted Boltzmann machine (RBM) is an undirected graphical model constructed for discrete or continuous random variables, with two layers, one hidden and one visible, and no conditional dependency within a layer. In recent years, RBMs have risen to prominence due to their connection to deep learning. By treating a hidden layer of one RBM as the visible layer in a second RBM, a deep architecture can be created. RBMs are thought to thereby have the ability to encode very complex and rich structures in data, making them attractive for supervised learning. However, the generative behavior of RBMs is largely unexplored. In this paper, we discuss the relationship between RBM parameter specification in the binary case and the tendency to undesirable model properties such as degeneracy, instability and uninterpretability. We also describe the difficulties that arise in likelihood-based and Bayes fitting of such (highly flexible) models, especially as Gibbs sampling (quasi-Bayes) methods are often advocated for the RBM model structure.
---

@@

where $\boldsymbol \theta \equiv (\theta_{11}, \dots, \theta_{1\nh}, \dots, \theta_{\nv 1}, \dots, \theta_{\nv \nh}, \theta_{v_1}, \dots, \theta_{v_\nv}, \theta_{h_1}, \dots, \theta_{h_\nh}) \in \mathbb{R}^{\nv + \nh + \nv*\nh}$ denotes the vector of model parameters and the denominator 
$$\gamma(\boldsymbol \theta) = \sum_{\boldsymbol x \in \mathcal{C}^{\nh+\nv}}\exp\left(\sum_{i = 1}^\nv \sum_{j=1}^\nh \theta_{ij} v_i h_j + \sum_{i = 1}^\nv\theta_{v_i} v_i + \sum_{j = 1}^\nh\theta_{h_j} h_j\right)
$$ 
is the normalizing function that ensures the probabilities (\ref{eqn:pmf}) sum to one. For $\boldsymbol x = (h_1, \dots, h_\nh, v_1, \dots, v_\nh) \in \mathcal{C}^{\nv + \nh}$ and 
\begin{align}
\boldsymbol t(\boldsymbol x) &= (h_1, \dots, h_\nh, v_1, \dots, v_\nv, v_1 h_1, \dots, v_\nv h_\nh) \in \mathcal{C}^{\nh + \nv + \nh*\nv}, \label{eqn:t}
\end{align}
let $\mathcal{T} = \{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{C}^{\nv + \nh}\} \subset \mathbb{R}^{\nv + \nh + \nv * \nh}$ be the set of possible values for the vector of variables needed to compute probabilities (\ref{eqn:pmf}) in the model and write $Q_{\boldsymbol \theta}(\boldsymbol x) = \sum\limits_{i = 1}^\nh\sum\limits_{j=1}^\nv \theta_{ij} h_i v_j + \sum\limits_{i = 1}^\nh\theta_{hi} h_i + \sum\limits_{j = 1}^\nv\theta_{vj} v_j$ for the "neg-potential" function. The RBM model is parameterized by $\boldsymbol \theta$ containing two types of parameters, main effects and interaction effects. The main effects parameters ($\{\theta_{v_i}, \theta_{h_j}\}_{\substack{i = 1, \dots, \nv,\\j = 1, \dots, \nh}}$) weight the values of the visible $v_i$ and hidden $h_j$ nodes in probabilities (\ref{eqn:pmf}) and the interaction effect parameters ($\theta_{ij}$) weight the values of the connections $v_i h_j$, or dependencies, between hidden and visible layers.

Due to the potential size of the model, the normalizing constant $\gamma(\boldsymbol \theta)$ can be practically impossible to calculate, making simple estimation of the model parameter vector problematic. A kind of Gibbs sampling can be tried (due to conditional architecture of the RBM, i.e. visibles given hiddens or vice versa). Specifically, the conditional independence of nodes in each layer (given those nodes in the other layer) allows for stepwise simulation of both hidden layers and model parameters (e.g., contrastive divergence of @hinton2002training or Bayes methods in Section \ref{model-fitting}).

@@

## Degeneracy, instability, and uninterpretability... Oh my!
The highly flexible nature of a RBM (having as it does $\nh + \nv + \nh*\nv$ parameters) creates at least three kinds of potential model impropriety that we will call *degeneracy*, *instability*, and *uninterpretability*. In this section we define these characteristics, consider how to detect them in a RBM, and point out relationships among them.

### Near-degeneracy
In Random Graph Model theory, *model degeneracy* means there is a disproportionate amount of probability placed on only a few elements of the sample space, $\mathcal{X}$, by the model [@handcock2003assessing]. For random graph models, $\mathcal{X}$ denotes all possible graphs that can be constructed from a set of nodes and an exponentially parameterized random graph model has a distribution of the form 
\begin{align*}
f_{\boldsymbol \theta} (\boldsymbol x) = \frac{\exp\left(\boldsymbol \theta^T \boldsymbol t(\boldsymbol x)\right)}{\gamma(\boldsymbol \theta)}, \boldsymbol x \in \mathcal{X},
\end{align*}
where $\boldsymbol \theta \in \Theta \subset \mathbb{R}^q$ is the model parameter, and $\boldsymbol t: \mathcal{X} \rightarrow \mathbb{R}^q$ is a vector of statistics based on the adjacency matrix of a graph. Here, as earlier, $\gamma(\boldsymbol \theta) = \sum_{\boldsymbol x \in \mathcal{X}} \exp\left(\boldsymbol \theta^T \boldsymbol t(\boldsymbol x)\right)$ is the normalizing function. Let $C$ denote the convex hull of the potential outcomes of sufficient statistics, $\{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{X}\}$. @handcock2003assessing classify an exponentially parametrized random graph model at $\boldsymbol \theta$ as *near-degenerate* if the mean value of the vector of sufficient statistics under $\boldsymbol \theta$, $\boldsymbol \mu(\boldsymbol \theta) = \text{E}_{\boldsymbol \theta}\boldsymbol t( \boldsymbol X)$, is close to the boundary of $C$. This makes sense because if a model is near-degenerate in the sense that only a small number of elements of the sample space $\mathcal{X}$ have positive probability, the expected value $\text{E}_{\boldsymbol \theta}\boldsymbol t( \boldsymbol X)$ is an average of that same small number of values of $t( \boldsymbol x)$ and can be expected to not be pulled deep into the interior of $C$.

@@

Considering exponential families of distributions, @schweinberger2011instability introduced a concept of model defficiency related to *instability*. Instability can be roughly thought of as excessive sensitivity in the model, where small changes in the components of potential data outcomes, $\boldsymbol x$, may lead to substantial changes in the probability function $f_{\boldsymbol \theta}(\boldsymbol x)$. To quantify "instability" more rigorously (particularly beyond the definition given by @schweinberger2011instability) it is useful to consider how RBM models might be expanded to incorporate more and more visibles. When increasing the size of RBM models, it becomes necessary to grow the number of model parameters (and in this process one may also arbitrarily expand the number of hidden variables used). To this end, let $\boldsymbol \theta_{\nv} \equiv (\theta_{v_1}, \dots, \theta_{v_\nv}, \theta_{h_1}, \dots, \theta_{h_\nh}, \theta_{11}, \dots, \theta_{\nv \nh}), \nv \ge 1$, denote an element of a sequence of RBM parameters indexed by the number $\nv$ of visibles $(V_1, \dots, V_\nv)$ and define a (scaled) extremal log-probability ratio of the RBM model at $\boldsymbol \theta_{\nv}$ as
\begin{align}
\frac{1}{\nv} \log \left[\frac{\max\limits_{(v_1, \dots, v_\nv) \in \mathcal{C}^\nv}P_{\boldsymbol \theta_\nv}(v_1, \dots, v_\nv)}{\min\limits_{(v_1, \dots, v_\nv) \in \mathcal{C}^\nv}P_{\boldsymbol \theta_\nv}(v_1, \dots, v_\nv)}\right] \equiv \frac{1}{\nv} \text{ELPR}(\boldsymbol \theta_\nv) \label{eqn:elpr}
\end{align}
where $P_{\boldsymbol \theta_\nv}(v_1, \dots, v_\nv) \propto \sum_{\{h_j\}\in \mathcal{C}}\exp\left(\sum_{i = 1}^\nv \theta_{v_i} v_i + \sum_{j = 1}^\nh \theta_{h_j} h_j + \sum_{i = 1}^\nv \sum_{j = 1}^\nh \theta_{ij} v_i h_j\right)$ is the RBM probability of observing outcome $(v_1, \dots, v_\nv)$ for the visible variables $(V_1, \dots, V_\nv)$ under parameter vector $\boldsymbol \theta_\nv$.

@@

S-unstable RBM model sequences are undesirable for several reasons. One is that, as mentioned above, small changes in data can lead to overly-sensitive changes in probability. Consider, for example, 
$$
\Delta(\boldsymbol \theta_\nv) \equiv \max \left\{\log \frac{P_{\boldsymbol \theta_\nv}(v_1, \dots, v_\nv)}{P_{\boldsymbol \theta_\nv}(v_1^*, \dots, v_\nv^*)} : (v_1, \dots, v_\nv) \& (v_1^*, \dots, v_\nv^*) \in \mathcal{C}^\nv \text{ differ in exactly one component}\right\},
$$
denoting the biggest log-probability ratio for a one component change in data outcomes (visibles) at a RBM parameter $\boldsymbol \theta_\nv$. We then have the following result.

@@

\begin{prop}
\label{prop:degen}
For an S-unstable RBM model sequence and any $0 < \epsilon < 1$, $P_{\boldsymbol \theta_\nv}\left((v_1, \dots, v_\nv) \in M_{\epsilon, \boldsymbol \theta_\nv}\right) \rightarrow 1$ as $\nv \rightarrow \infty$.
\end{prop}

In other words, S-unstable RBM model sequences are guaranteed to stack up all probability on an arbitrarily narrow set of outcomes for visibles. Proofs of Propositions \ref{prop:instab}-\ref{prop:degen} appear in @kaplan2016note. These findings also have counterparts in results in @schweinberger2011instability, but unlike results there, we do not limit consideration to exponential family forms with a fixed number of parameters.

### Uninterpretability
For spatial Markov models, @kaiser2007statistical defines a measure of model impropriety he calls *uninterpretability*, which is characterized by dependence parameters in a model being so extreme that marginal mean-structures fail to hold as anticipated by a model statement. We adapt this notion to RBM models as follows. Note that in a RBM, the parameters $\theta_{v_1}, \dots, \theta_{v_\nv}$ and $\theta_{h_1}, \dots, \theta_{h_\nh}$ are naturally associated with main effects of visible and hidden variables and can be interpreted as (logit functions of) means for variables $V_1, \dots, V_\nv, H_1, \dots, H_\nh$ in a model with no interaction parameters, $\theta_{ij} = 0, i = 1, \dots, \nv, j = 1, \dots, \nh$. That is, with no interaction parameters, we have from (\ref{eqn:pmf}) that
$$
P_\theta(V_i=1) \propto e^{\theta_{v_i}} \quad \text{ and } \quad P_\theta(H_j=1) \propto e^{\theta_{h_j}}, \quad  i=1,\ldots,\nv,j=1,\ldots,\nh
$$
so that, for example, $\text{logit}(P_\theta(V_i=1))=\theta_{v_i}$ (or $2 \theta_{v_i}$) under the coding $\mathcal{C}=\{0,1\}$ (or $\{-1,1\}$). Hence, these main effect parameters have a clear interpretation under an independence model (one with $\theta_{ij} = 0$) but this interpretation can break down as interaction parameters increase in magnitude relative to the size of the main effects. In such cases, the main effect parameters $\theta_{v_1}$ and $\theta_{h_j}$ are no longer interpretable as originally intended in the models (statements of marginal means) and the dependence parameters are so large as to dominate the entire model probability structure (also destroying the model interpretation of dependence as local conditional modifications of an overall marginal mean structure). To assess which parameter values $\boldsymbol \theta$ may cause difficulties in interpretation, we use the difference $\text{E}\left[\boldsymbol X | \boldsymbol \theta\right] - \text{E}\left[\boldsymbol X | \boldsymbol \theta^* \right]$ between two model expectations: E$\left[\boldsymbol X | \boldsymbol \theta\right]$ at $\boldsymbol \theta$ and expectations E$\left[\boldsymbol X | \boldsymbol \theta^* \right]$ where $\boldsymbol \theta^*$ matches $\boldsymbol \theta$ for all main effects but otherwise has $\theta_{ij} = 0$ for $i = 1, \dots, \nv, j = 1, \dots, \nh$. Hence, $\boldsymbol \theta^*$ and $\boldsymbol \theta$ have the same main effects but $\boldsymbol \theta^*$ has $0$ dependence parameters. Uninterpretability is then avoided at a parametric specification $\boldsymbol \theta$ if the model expected value at $\boldsymbol \theta$ is not very different from the corresponding model expectation under independence. Using this, it is possible to investigate what parametric conditions lead to uninterpretability in a model versus those that guarantee interpretable models. If $||\text{E}\left[\boldsymbol X | \boldsymbol \theta\right] - \text{E}\left[\boldsymbol X | \boldsymbol \theta^*\right]||$ is large, then the RBM model with parameter vector $\boldsymbol \theta$ is said to be uninterpetable. The quantities to compare in the RBM case are

@@

We next explore and numerically explain the relationship between values of $\boldsymbol \theta$ and the three notions of model impropriety, near-degeneracy, instability, and uninterpretability, for RBM models of varying sizes.

@@

Figure \ref{fig:uninterp} shows that the absolute difference between coordinates of the vector of model expectations, E$\left[\boldsymbol X | \boldsymbol \theta\right]$ and corresponding expectations E$\left[\boldsymbol X | \boldsymbol \theta^*\right ]$ given independence grow for the toy RBM as the values of $\boldsymbol \theta$ are farther from zero, especially for large magnitudes of $\theta_{11}$. This is a third indication that parameter vectors of large magnitude lead to model impropriety in a RBM.

@@

In addition to the argument that the $-1/1$ data encoding mitigates some perceived prevalence of near-degeneracy it also has the benefit of providing a guaranteed-to-be non-degenerate model at $\boldsymbol \theta = \boldsymbol 0 \in \mathbb{R}^{\nh + \nv + \nh*\nv}$, where the zero vector that serves as the natural center of the parameter space and induces the simplest possible model properties for the RBM (i.e., at $\boldsymbol \theta = \boldsymbol 0$, all variables are independent and visible variables are independent and uniformly distributed on $\{-1,1\}^\nv$). The proof of this and further exploration of the equivalence of the $\boldsymbol \theta$ parameterization of the RBM model class and parameterization by $\boldsymbol \mu(\boldsymbol \theta)$ is in the on-line supplementary materials. Hence while from some computing perspectives $0/1$ coding might seem most natural, the $-1/1$ coding is far more convenient and interpretable from the point of view of statistical modeling, where it makes parameters simply interpreted in terms of symmetrically defined main effects and interactions. In light of all of these matters we will henceforth employ the $-1/1$ coding.

@@

To explore the impact of RBM parameter vector $\boldsymbol \theta$ magnitude on near-degeneracy, instability, and uninterpretability, we consider models of small size. For $\nh, \nv \in \{1, \dots, 4\}$, we sample `r dim(res$samp[[1]])[1]` values of $\boldsymbol \theta$ with various magnitudes (details to follow). For each set of parameters we then calculate metrics of model impropriety introduced in Section \ref{degeneracy-instability-and-uninterpretability-oh-my} based on $\boldsymbol \mu(\boldsymbol \theta)$, $\text{ELPR}(\boldsymbol \theta)/\nv$, and the absolute coordinates of $\text{E}\left[\boldsymbol X | \boldsymbol \theta\right] -   \text{E}\left[\boldsymbol X | \boldsymbol \theta^* \right]$. In the case of near-degeneracy, we classify each model as near-degenerate or "viable" based on the distance of $\boldsymbol \mu(\boldsymbol \theta)$ from the boundary of the convex hull of $\mathcal{T}$ and look at the fraction of models that are "near-degenerate," meaning they are within a small distance $\epsilon > 0$ of the boundary of the convex hull. We define "small" through a rough estimation of the volume of the hull for each model size. We pick $\epsilon_0 = 0.05$ for $\nh=\nv=1$ and then, for every other $\nh$ and $\nv$, set $m=\nh+\nv+\nv*\nh$ and pick $\epsilon$ so that $1-(1-2\epsilon_0)^3 = 1 - (1-2\epsilon)^m$. In this way, we roughly scale the volume of the "small distance" to the boundary of the convex hull to be equivalent across model dimensions.

In our numerical experiment, we split $\boldsymbol \theta = (\boldsymbol \theta_{main}, \boldsymbol \theta_{interaction})$ into $\boldsymbol \theta_{main}$ and $\boldsymbol \theta_{interaction}$, in reference to which variables in the probability function the parameters correspond (whether they multiply a $v_i$ or a $h_j$ or they multiply a $v_i h_j$), and allow the two types of terms to have varying average magnitudes, $||\boldsymbol \theta_{main} || /(\nh+\nv)$ and $||\boldsymbol \theta_{interaction} || /(\nh*\nv)$. These average magnitudes vary on a grid between `r min(plot_dat_grid$r1/(plot_dat_grid$H + plot_dat_grid$V))` and `r max(plot_dat_grid$r1/(plot_dat_grid$H + plot_dat_grid$V))` with `r length(unique(plot_dat_grid$r1/(plot_dat_grid$H + plot_dat_grid$V)))` breaks, yielding `r length(unique(plot_dat_grid$r1/(plot_dat_grid$H + plot_dat_grid$V)))^2` grid points. (By looking at the average magnitudes, we are able to later consider the potential benefit of shrinking each parameter value $\theta_i$ towards zero in a Bayesian fitting technique.) At each point in the grid, `r dim(res$samp[[1]])[1]` vectors ($\boldsymbol \theta_{main}$) are sampled uniformly on a sphere with radius corresponding to the first coordinate in the grid and `r dim(res$samp[[1]])[1]` vectors ($\boldsymbol \theta_{interction}$) are sampled uniformly on a sphere with radius corresponding to the second coordinate in the grid via sums of squared and scaled iid Normal$(0, 1)$ variables. These vectors are then paired to create `r dim(res$samp[[1]])[1]` values of $\boldsymbol \theta$ with magnitudes at each point in the grid.
